The Spatial and Temporal Domains of Modern Ecology

### 2016-08-25 paper/SI.tex
Final draft with Erle's edits. Still left: full review bibliography
Updated ms to partially match Nature EcoEvo format, mainly with fuller Nature-style abstract.

\renewcommand\refname{References and Notes}

@@

Each observer first reviewed the papers in the calibration set, and then commenced reviewing papers in their individual random draws, beginning at the top of the list and then proceeding until at least 20 eligible papers describing ecological observations were reviewed. In cases where the reviewed papers used observations that were described in another publication, we reviewed those source papers in order to extract the observational dimensions. We excluded papers that were opinion or perspectives pieces (unless they presented or used existing observational data), theoretical studies based on generated data, or those which were entirely based on experimental manipulations. We left out the latter category because our intent was to evaluate the domains for observations of natural systems, and we wanted to avoid the bias that would be imposed by the relatively narrow spatial and temporal scales of experiments \cite{kareiva_spatial_1988,tilman_global_2011}.

@@

\noindent\underline{\textbf{General:}}
\def\Hitem{\item [Q\stepcounter{enumi}\arabic{enumi}.]} 
\def\Hsubitem{\item [Q\arabic{enumi}.\arabic{enumi}.]} 
\vspace{-0.5cm}
\begin{enumerate}
  \Hitem \emph{\textbf{What are the general inclusion/exclusion criteria for studies?}} Studies should be excluded from this analysis if they are: 1) opinion/perspectives pieces; 2) model-only studies, particularly theoretical models, which are not developed or tested against observed data; 3) if they are experimental manipulations (but if there is a study that has a mix of observational and experimental, record the observational treatments and exclude the manipulated treatments).

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we estimated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in a study with an unclear or missing scale value, which we used to assess a dimension-specific estimate of uncertainty.

@@

\bibliographystyle{Science}

% Following is a new environment, {scilastnote}, that's defined in the
% preamble and that allows authors to add a reference at the end of the
% list that's not signaled in the text; such references are used in
% *Science* for acknowledgments of funding, help, etc.

%\begin{scilastnote}
%\item We've included in the template file \texttt{scifile.tex} a new
%environment, \texttt{\{scilastnote\}}, that generates a numbered final
%citation without a corresponding signal in the text.  This environment
%can be used to generate a final numbered reference containing
%acknowledgments, sources of funding, and the like, per {\it Science\/}
%style.
%\end{scilastnote}

% For your review copy (i.e., the file you initially send in for
% evaluation), you can use the {figure} environment and the
% \includegraphics command to stream your figures into the text, placing
% all figures at the end.  For the final, revised manuscript for
% acceptance and production, however, PostScript or other graphics
% should not be streamed into your compliled file.  Instead, set
% captions as simple paragraphs (with a \noindent tag), setting them
% off from the rest of the text with a \clearpage as shown  below, and
% submit figures as separate files according to the Art Department's
% instructions.

\clearpage

>>>>

%\renewcommand\refname{References and Notes}

@@

Each observer first reviewed the papers in the calibration set, and then commenced reviewing papers in their individual random draws, beginning at the top of the list and then proceeding until at least 20 eligible papers describing ecological observations were reviewed. In cases where the reviewed papers used observations that were described in another publication, we reviewed those source papers in order to extract the observational dimensions. We excluded papers that were opinion or perspectives pieces (unless they presented or used existing observational data), theoretical studies based on generated data, or those which were entirely based on experimental manipulations. We left out the latter category because our intent was to evaluate the domains for observations of natural systems, and we wanted to avoid the bias that would be imposed by the relatively narrow spatial and temporal scales of experiments \cite{kareiva_spatial_1988,tilman_global_2011}.  A bibliography of the reviewed papers follows the References and Notes section below.

@@

\noindent\underline{\textbf{General:}}
\def\Hitem{\item [Q\stepcounter{enumi}\arabic{enumi}.]} 
\def\Hsubitem{\item [Q\arabic{enumi}.\arabic{enumi}.]} 
\vspace{-0.5cm}
\begin{enumerate}
  \Hitem \emph{\textbf{What are the general inclusion/exclusion criteria for studies?}} Studies should be excluded from this analysis if they are: 1) opinion/perspectives pieces; 2) book reviews; 3) model-only studies, particularly theoretical models, which are not developed or tested against observed data; 4) if they are experimental manipulations (but if there is a study that has a mix of observational and experimental, record the observational treatments and exclude the manipulated treatments).

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we estimated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in a study with an unclear or missing scale value.

@@

\bibliographystyle{naturemag}

### 2016-08-25 paper/main.tex
Updated ms to partially match Nature EcoEvo format, mainly with fuller Nature-style abstract.

\author
{Lyndon Estes$^{\ast1, 2}$, Labeeb Ahmed$^{3}$, Kelly Caylor$^{2}$, Jason Chang$^{3}$, \\
Jonathan Choi$^{4}$, Erle Ellis$^{3}$, Paul R. Elsen$^{1}$, and Tim Treuer$^{4}$ \\
\\
\normalsize{$^{1}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{2}$Civil and Environmental Engineering, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  lestes@princeton.edu.}
}
% Include the date command, but leave its argument blank.

@@

\begin{sciabstract}
Observational scales play a critical role in shaping understanding of ecological pattern and process. To gain insight into the spatial and temporal domains being addressed by modern ecological science, we conducted a meta-analysis of papers published in 30 ecological journals between 2004 and 2014. We estimated two spatial dimensions (resolution and extent) and two temporal dimensions (interval and duration) of 371 ecological observations reported within 140 papers (out of 364 randomly selected titles). Our analysis revealed that ecology largely remains a field-based discipline that primarily collects observations within narrow spatial and temporal domains. Ecologists still make limited use of methods for observing larger-scale patterns and processes (e.g. remote sensing, used for 6\% of observations), but make greater use of those which permit high frequency or temporally continuous \emph{in situ} measurements (12\% of observations). Resolution, extent, and duration were not clearly reported for the majority of observations, which, together with the narrow spatio-temporal ranges of observed scales, suggests a limited ability to generalize the findings of many studies.

%  An example of the style is the special\texttt{\{sciabstract\}} environment used to set up the abstract you see here.
\end{sciabstract}

\linenumbers
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989-1}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is the \emph{sine qua non} for understanding ecological phenomena \cite{levin_problem_1992}. Awareness of the importance of scale has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad range of scales \cite{schneider_rise_2001,chave_problem_2013}.

Given the centrality of scale to understanding ecology, and the growing ability to study ecological phenomena across a broad range of spatial and temporal scales, it is important to assess the scales at which ecological research is presently being conducted. To gain insight into this question, we quantified the spatial and temporal domains of empirical observations that were reported in a representative sample of studies published between 2004-2014 in the top 30 ecological journals (by 2014 impact factor). Empirical observations provide the necessary means for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, thus it stands to reason that the temporal and spatial distributions of ecological observations may be indicative of modern ecology's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

We characterized observational domains along two key spatial dimensions, resolution (grain) and extent, and their temporal corollaries, sampling interval and total temporal duration. We calculated the actual, as opposed to representative (SI), scales for these four dimensions from 371 discrete observations (defined here as data collected from non-experimentally manipulated, or ``natural'' \cite{tilman_ecological_1989}, systems) reported within a 140 paper subset of 364 randomly selected articles (from 42,918 total). To account for uncertainty in the estimation of observational dimensions due to 1) unclear methodological description in the reviewed papers, and 2) observer interpretation, we conducted a resampling analysis (n=1000) in which scale values were randomly perturbed within the bounds of estimated inter-observer variation (SI). We constructed histograms for each dimension from the mean of the perturbed ensembles, and estimated 95\% confidence intervals for each histogram bin (Fig. 1). We constructed  kernel density estimates from the full resampled ensemble in order to assess observational distributions within different juxtapositions of the four space-time dimensions (Fig. 2).

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
Despite these caveats, our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989-1}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989-1}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989-1}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.

@@

%These findings suggest that ecological understanding is shaped by studies of relatively fine-scaled patterns, and because they are small in extent may be less generalizable \cite{wiens_spatial_1989-1}.

@@

%Based on the observational domains recorded here, it appears that ecologists are still primarily collecting data at very fine spatial scales and over very narrow extents. Although the temporal duration of most ecological observations is also narrow, and a sizable number of observations are not repeated, greater effort is being devoted to capturing temporal dynamics. These findings suggest that ecological understanding is shaped by studies of relatively fine-scaled patterns, and because they are small in extent may be less generalizable \cite{wiens_spatial_1989-1}.

@@

%\paragraph*{Displayed math.}

\bibliography{/Users/lestes/Dropbox/publications/fullbib}

\bibliographystyle{Science}

@@

\begin{scilastnote}
\item This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).
\end{scilastnote}

>>>>

\author
{Lyndon Estes$^{\ast1, 2}$, Labeeb Ahmed$^{3}$, Kelly Caylor$^{2}$, Jason Chang$^{3}$, \\
Jonathan J. Choi$^{4}$, Erle Ellis$^{3}$, Paul R. Elsen$^{1}$, and Tim Treuer$^{4}$ \\
\\
\normalsize{$^{1}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{2}$Civil and Environmental Engineering, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  lestes@princeton.edu.}
}
% Include the date command, but leave its argument blank.

@@

%\begin{sciabstract}
\textbf{Observational scale shapes our understanding of ecosystems \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}. Ecological patterns emerge at temporal and spatial scales that may be coarser or finer than the processes that shape them, thus investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992}. Given the centrality of scale to ecology, and the growing ability to observe ecological phenomena beyond a field perspective, it is important to understand the scales at which modern ecological research is conducted. To reveal these, we conducted a meta-analysis of recent (2004-2014) ecological research, in which we estimated the spatial resolution and extent, and temporal interval and duration, of reported ecological observations. Here we show that ecology remains a largely field-based discipline that makes observations within generally narrow spatial and temporal domains, despite the well-established literature on the importance of scale \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}. Ecologists still make limited use of remote sensing (6\% of observations), which is necessary for observing larger-scale patterns and processes \cite{turner_remote_2003,pettorelli_satellite_2014}, but make slightly greater use of technologies that capture high frequency or temporally continuous point-based measurements (12\% of observations). Resolution, extent, and duration were not clearly reported in most reviewed studies. This indicates inattentiveness to scale, which, together with the narrow spatio-temporal domains of observations, suggests a limited ability to generalize the knowledge gained in many ecological studies, which is critical for developing an ability to predict ecological responses in an era of increasing global change \cite{levin_problem_1992}. An emerging wave of observing and analytical technologies (e.g. nano-satellites, unmanned systems, open image archives, and high-power processing platforms) now makes it dramatically easier and cheaper to collect high-quality, ecologically meaningful measurements at scales inaccessible to field scientists. Combined with a greater attention to scale in research design, these advances can facilitate the spread of ecological research to sparsely observed scales, while improving the transferability of ecological knowledge.}

%Observational scales play a critical role in shaping our understanding of ecological pattern and process. To gain insight into the spatial and temporal domains being addressed by modern ecological science, we conducted a meta-analysis of papers published in 30 ecological journals between 2004 and 2014. We estimated two spatial dimensions (resolution and extent) and two temporal dimensions (interval and duration) of 371 ecological observations reported within 140 papers (out of 346 randomly selected titles). Our analysis revealed that ecology largely remains a field-based discipline that primarily collects observations within narrow spatial and temporal domains. Ecologists still make limited use of methods for observing larger-scale patterns and processes (e.g. remote sensing, used for 6\% of observations), but make greater use of those which permit high frequency or temporally continuous \emph{in situ} measurements (12\% of observations). Resolution, extent, and duration were not clearly reported for the majority of observations, which, together with the narrow spatio-temporal ranges of observed scales, suggests a limited ability to generalize the findings of many studies.

%  An example of the style is the special\texttt{\{sciabstract\}} environment used to set up the abstract you see here.
%\end{sciabstract}

\linenumbers
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is the \emph{sine qua non} for understanding ecological phenomena \cite{levin_problem_1992}. Awareness of the importance of scale has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad range of scales \cite{schneider_rise_2001,chave_problem_2013}.

Given the centrality of scale to understanding ecology, and the growing ability to study ecological phenomena across a broad range of spatial and temporal scales, it is important to assess the scales at which ecological research is presently being conducted. To gain insight into this question, we quantified the spatial and temporal domains of empirical observations that were reported in a representative sample of studies published between 2004-2014 in the top 30 ecological journals (by impact factor). Empirical observations provide the necessary means for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, thus it stands to reason that the temporal and spatial distributions of ecological observations may be indicative of modern ecology's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

We characterized observational domains along two key spatial dimensions, resolution (grain) and extent, and their temporal corollaries, sampling interval and total temporal duration. We calculated the actual, as opposed to representative (SI), scales for these four dimensions from 371 discrete observations (defined here as data collected from non-experimentally manipulated, or ``natural'' \cite{tilman_ecological_1989}, systems) reported within a 140 paper subset of 346 randomly selected articles (from 42,918 total). An additional 56 papers that were cited as the source of observations in the selected papers were also reviewed. To account for uncertainty in the estimation of observational dimensions due to 1) unclear methodological description in the reviewed papers, and 2) observer interpretation, we conducted a resampling analysis (n=1000) in which scale values were randomly perturbed within the bounds of estimated inter-observer variation (SI). We constructed histograms for each dimension from the mean of the perturbed ensembles, and estimated 95\% confidence intervals for each histogram bin (Fig. 1). We constructed  kernel density estimates from the full resampled ensemble in order to assess observational distributions within different juxtapositions of the four space-time dimensions (Fig. 2).

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
Despite these caveats, our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.

@@

%These findings suggest that ecological understanding is shaped by studies of relatively fine-scaled patterns, and because they are small in extent may be less generalizable \cite{wiens_spatial_1989}.

@@

%Based on the observational domains recorded here, it appears that ecologists are still primarily collecting data at very fine spatial scales and over very narrow extents. Although then temporal duration of most ecological observations is also narrow, and a sizable number of observations are not repeated, greater effort is being devoted to capturing temporal dynamics. These findings suggest that ecological understanding is shaped by studies of relatively fine-scaled patterns, and because they are small in extent may be less generalizable \cite{wiens_spatial_1989}.

@@

%\paragraph*{Displayed math.} 
\bibliography{/Users/lestes/Dropbox/publications/fullbib}
\bibliographystyle{naturemag}

\noindent \textbf{Acknowledgements} This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).

@@

%\begin{scilastnote}
%\item This work was supported by funds from the Princeton Environmental Institute Grand Challenges program and the NASA New Investigator Program (NNX15AC64G). Erle Ellis, Jason Chang, and Labeeb Ahmed of the GLOBE Project (http://globe.umbc.edu) were supported by the U.S. National Science Foundation (1125210).
%\end{scilastnote}
